# Control flow

## For loops

A for loop can be inefficient in R if you are growing a data structure (e.g. dataframe) because R will recreate the object in memory at each iteration. However, if you create a vector of the correct size (or of a size that you know will be more than enough) outside of the for loop, then fill this vector in your loop, it is more efficient. In this case R will create a structure with enough slots prior to the loop, instead of asking for more space at each iteration. A good reference on this isse is [this](https://kbroman.wordpress.com/2013/04/02/apply-vs-for/) blog

Objects created in a for loop are global. This is as opposed to objects created within a function, which are local.

### Creating empty data structures

Atomic vector can be created with any of these functions. The only argument they take is the length of the vector.

```{r}
str(logical(10))
```

Empty logical vectors are filled with FALSE by default.

```{r}
str(numeric(10))
str(double(10))
str(integer(10))
```

Empty number vectors are filled with 0's by default.

```{r}
str(character(10))
```

Empty mumeric vectors are filled with empty strings by default.

You can also create a vector using `vector()`:

```{r}
str(vector(mode = "character", length = 10))
```

A list can be created using this function:

```{r}
str(vector(mode = "list", length = 10))
```

An empty list is filled with NULLs by default.

An empty matrix can be created thus:

```{r}
str(
  matrix(nrow=2, ncol = 2)
)
```

An empty matrix is filled with NAs by default.

A dataframe can be created in two ways:

```{r}
str(
  data.frame(col1 = numeric(10), col2 = character(10), 
             stringsAsFactors = FALSE)
)
```

It is important to add `stringsAsFactors = FALSE`, otherwise character vectors will be automatically converted into factors.

```{r}
str(
  data.frame(matrix(nrow=2, ncol = 2))
)
```

Growing a dataframe efficiently in R is difficult. The best option may be to create vectors in the loop and then join the vectors.

```{r, eval=FALSE}
x <- numeric(n)
y <- character(n)
  
for (i in 1:n) {
  x[i] <- i
  y[i] <- i
}

data.frame(x, y, stringsAsFactors=FALSE)
```

See [this](https://stackoverflow.com/questions/11486369/growing-a-data-frame-in-a-memory-efficient-manner) question on SO for more. (There is also a data.table method - see [this](https://stackoverflow.com/questions/20689650/how-to-append-rows-to-an-r-data-frame/38052208#38052208) SO question and section 888 for more.)




