# Functions

Useful definition:

* parameter is what is given in the function declaration/definition.
* argument is what is passed when called the function.

## Source code

You can view the source code of functions in R.

For 'Generic' S3 objects (in RStudio): 

* Type in the function name with no '()'.
* Press F2 when your cursor is over the function 

`methods()` will list all the methods for a particular generic function or class.

For S4 objects just typing in the function name does not give you the source code. It will give some other useful information - e.g. if it says 'standardGeneric' means tells you that it is a S4 function.

`showMethods()` - tells you the available methods for that function (e.g. if a function performs a slightly different thing depending on the class of object given to it).
`getMethod()` - can then be used to see the source code for one of the available methods in showMethods.

(see [this](https://stackoverflow.com/questions/19226816/how-can-i-view-the-source-code-for-a-function) SO question for more).

## Writing functions

Standard syntax:

```{r, eval=FALSE}
function_name <- function(param1, param2) {
  result <-  param1 - param2
  return(result)
}
```

### Evaluation

For column names the `$` operator will not work if passed in as argument.

```{r, eval=FALSE}
foo <- function(df,col_name, col1,col2){
      df$col_name <- df$col1 + df$col2
      df
}
```

Above function does not work - as the function will look for a column called 'col1' instead of evaluating expression col1 (i.e. the column name assigned to the argument col1).

The easiest solution is to use `[[`:


```{r, eval=FALSE}
new_column1 <- function(df,col_name,col1,col2){
    #Create new column col_name as sum of col1 and col2
    df[[col_name]] <- df[[col1]] + df[[col2]]
    df
}
```

Note when using `[[` , you need to quote the column names, thus the argument should be given as a string.

See [this](https://stackoverflow.com/questions/2641653/pass-a-data-frame-column-name-to-a-function) question on SO for more.

### Checking inputs

`assert_that()` from the asserthat package lets you check inputs and provdie error messages.

```{r, eval=FALSE}
assert_that(is.logical(reps), length(reps) == 1,     
              msg = "error message")  
```

The function will stop and the error message displayed if EITHER of the two conditions are not met.

assert_that will automatically format your error messages so that even if you put a newline in your code, it will print well for the user.

If you are not using `assert_that()`, you can use `strwrap()` to wrap long error messages, so they do not overflow 80 characters in your code but when printed it will fit screen of the user. 

See [this](https://stackoverflow.com/questions/45693010/how-do-you-format-multiline-r-package-messages) question on SO for more.

## Messages

There are two types of output messages:

* Standard output - `print() and cat() sends to stdout
* Standard error - these are able to be suppressed using `supressWarnings()`. Useful if the user is using knitr/sweave.

When using a R script on the command line - you can redirect standard out and standard error (recall 1 is standard out, 2 is standard error). 

- **Standard out**
    - **print()**
        * Need to use with paste to concatenate terms.
        * Prints [x] at the start, where x indicates the element of the term.
    - **cat()**
        * Allows concatenation and has the sep argument.
        * Need to either for new lines with ‘\n' or `fill = TRUE` (uses default print width).
- **Standard error**
    - **message()**   
        * Output is usually red.
        * Allows concatenation but no `sep` argument.
    - **warning()**
        * 'Warning message' line preappended to it.
        * Do not use lightly - sometimes treated as errors during testing.
    - **stop()**
        * Stops the task at hand.
        * 'Error:' printed at the start of the line.

### Optional arguments

Use `missing()`. Below, `y` is an optional argument.

```{r, eval=FALSE}
fooBar <- function(x,y){
    if(missing(y)) {
        x
    } else {
        x + y
    }
}
```

## Functions with dplyr

References: [dplyr documentation](https://dplyr.tidyverse.org/articles/programming.html), [SO](https://stackoverflow.com/questions/27197617/filter-data-frame-by-character-column-name-in-dplyr)

### Non standard evaluation

Non standard evalation is a catch-all term that means dplyr functions do not follow the usual R rules of evaluation. There are two main benefits of this:

1. Saving on typing. You do not have to type df name again:

`filter(df, x == 1, y == 2, z == 3)` instead of `df[df$x == 1 & df$y ==2 & df$z == 3, ]`.

2. dplyr can choose to compute results in a different way to base R. This is important for database backends because dplyr itself doesn't do any work, but instead generates the SQL that tells the database what to do.

The drawbacks:

1. dplyr arguments are not 'referentially transparent'. You cannot replace a value with a seemingly equivalence object that you've defined elsewhere.

```{r}
# define df
df <- tibble(x = 1:3, y = 3:1)
```

This code: 

```{r}
filter(df, x == 1)
```

is NOT the same as this code:

```{r, error=TRUE}
my_var <- x
filter(df, my_var == 1)
```

Nor this code:

```{r, error=TRUE}
my_var <- "x"
filter(df, my_var == 1)
```

This makes creating functions with arguments that need to be used with dplyr functions.

2. The code is ambiguous. Depending on what variables are defined where, `filter(df, x == y)` could be equivalent to any of:  

```{r, eval=FALSE}
df[df$x == df$y, ]
df[df$x == y, ]
df[x == df$y, ]
df[x == y, ]
```


Analogous evaluation problem:

```{r}
greet <- function(name) {
  "How do you do, name?"
}
greet("Hadley")
```

The quotes in:  `"How do you do, name?"` mean that it doesn't interpret what you've typed. It just stores it in a string. 

### Writing dplyr functions

dplyr does not do anything special with the 1st argument, the data object.

This works fine:

```{r, eval=FALSE}
mutate_y <- function(df) {
  mutate(df, y = a + x)
}
```

Though, it can silently fail if one of the variables is not present in the data frame but present in the global environment.

For example, if you function added two columns: 'y = column a + column x' but there is no column `a` but in the global environment is a variable `a` (`a <- 3`) - 

```{r, eval=FALSE}
mutate_y <- function(df) {
  mutate(df, y = a + x)
}
```

Will add the variable `a`, to the column x.

To prevent this, you can use this more explicit code:

```{r, eval=FALSE}
mutate_y <- function(df) {
  mutate(df, y = .data$a + .data$x)
}
```

`.data` specifies that you want to use the columns of the dataframe.

If you want one of the arguments to be a variable name or an expression (e.g. x+y) - you need to handle this specially.

For example, if you want to write a function that does this:

```{r}
my_summarise <- function(df, group_var) {
  df %>%
    group_by(group_var) %>%
    summarise(a = mean(Petal.Length))
}
```

```{r, error=TRUE}
my_summarise(iris, Species)
```

An error is given. This is because dplyr 'quotes' the input to `group_by()`. It does not evaluate what you give it (here `group_var`).

To fix this we need to do 2 things:
We need to quote the input ourselves. We use `quo()` which returns a quosure, which is a special type of formula. This has 2 uses:

If we just give an unquoted argument, R will use it as a variable. It will then tell you it can't find this variable. 

`quo()` function quotes its input. We cannot just use '' as it would give us a string. `quo()` captures the expression and its environment.

We then need to tell the dplyr function to not quote its input as we have already done it. We can do this with `!!`.

```{r}
my_summarise <- function(df, group_var) {
  df %>%
    group_by(!! group_var) %>%
    summarise(a = mean(Petal.Length))
}
```

```{r}
my_summarise(iris, quo(Species))
```

But what if we did not want to give `quo(group_var)`? We can try this:

```{r}
my_summarise <- function(df, group_var) {
  
  quo_group_var <- quo(group_var)
  print(quo_group_var)

  df %>%
    group_by(!! quo_group_var) %>%
    summarise(a = mean(Petal.Length))
}
```

```{r, error=TRUE}
my_summarise(iris, Species)
```


This error occurs because `quo(group_var)` actually quotes 'group_var'. What we want is for it to substitute the value that the user supplied.

Instead we can use `enquo()`. This function looks at the argument, see what the actual input was and quotes that (instead of the parameter name).

```{r}
my_summarise <- function(df, group_var) {
  group_var <- enquo(group_var)
  print(group_var)

  df %>%
    group_by(!! group_var) %>%
    summarise(a = mean(Petal.Length))
}
```

Other options are:

```{r, eval=FALSE}
col1 <- "Sepal.Length"
iris %>%
  filter(UQ(as.name(col1)) > 5.5)


col1 <- "Sepal.Length"
iris %>%
  filter(!!(as.name(col1)) > 5.5)
```

The `!!` function works (in a dplyr pipe) even if you do not have rlang (the package the function comes from) loaded.

## ggplot

ggplot also uses non-standard evaluation - you can refer to columns in your data without specifying the dataframe first. Again, this means that you have to do something special if you want to use ggplot in a function and refer to columns provided as arguments to your function.

Use `aes_string` instead of `aes()`: 

* `aes()` does complicated munging to get the names of the variables.
* `aes_string()` works directly with strings.

`aes_string(x = "mpg", y = "wt")` is equal to `aes(x = mpg, y = wt)`.

```{r, eval=FALSE}
f <- function(column) {
    ...
    ggplot(df, aes_string(x = "col1", y = "col2", colour = column, 
                          group = column) )
}
```

Column name arguments passed to your function must be quoted.

You can also use `get()`:

```{r, eval=FALSE}
ggplot(d, aes(x = name, y = rate, colour = get(column),
              group = get(column)))
```

See [this](https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot) question on SO for more.











      



































